---
author: "Ning Yilin"
date: "Last update on `r Sys.Date()`"
title: "General Tips"
---

```{r, include = FALSE}
# I personally prefer this style because it provides a floating table of 
# content:
'
output:
  html_document:
    theme: flatly
    highlight: zenburn
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
'
# But the knitted HTML file would be too large for Github to display. If you 
# are interested you can replace the current `output` chunk in the YAML code 
# above and try yourself.
```

<a name="top"></a>
<a style="display:scroll; position:fixed; bottom:5px; right:5px;" href="#top"><font size="-1">[top]</font></a>
<a style="display:scroll; position:fixed; bottom:5px; left:5px;" href="https://github.com/nyilin/My-R-help-sheets"><font size="-1">[Back to index]</font></a>

```{r, include = FALSE}
# global options
library(knitr)
opts_chunk$set(collapse = TRUE, echo = TRUE, eval = TRUE, cache = FALSE)
```

# About R Markdown

This document is written using R Markdown and `knitr`. For more details on how 
to write R Markdown files, see 
[this](http://rmarkdown.rstudio.com/html_document_format.html) website.

The following two lines, taken from the 
[webpage for package `Armadillo`](http://arma.sourceforge.net/docs.html), 
add a floating back-to-top buttom at the bottom right of the page.

```{html}
<a name="top"></a>
<a style="display:scroll; position:fixed; bottom:5px; right:5px;" href="#top"><font size="-1">[top]</font></a>
```

# Use RStudio

RStudio is a powerful IDE for `R` language that could be used to build up R 
code, packages and projects (with version control supported by `Git` and `SVN`), 
and produce `pdf`, `docx` or `html` reports with additional packages. 
**Use RStudio whenever possible.**

# Style guide

A good code style makes code maintainance and sharing much more easily. Refer to 
[this](https://google.github.io/styleguide/Rguide.xml) website for details.

Some recommended practive:

- Be consistent
- Put empty space around math operators
- Use `<-` instead of `=` to assign value
- Put an empty space between `if`, `for` and `while` and the openning bracket
- Put an empty space before the openning curly bracket and after the closing
curly bracked

# Resources

## Quick-R

[This website](http://www.statmethods.net/index.html) illustrates basic R 
functions and statistics, including input (file formats, data structure, missing 
data, date value), output, data type conversion.

[This page](http://www.statmethods.net/advgraphs/parameters.html) in the website 
shows commonly used `par()` arguments including `pch`, `lty` and `text`.

## Advanced R

[This book](http://adv-r.had.co.nz/) is written by 
[Hadley Wickham](http://hadley.nz/), the author of many popular R packages 
including `devtools`, `roxygen2`, `ggplot` and `dplyr`. The book provides an 
in-depth review of R as a programing language.

# Basic Commands and Keyboard Shortcuts

## Install and remove packages

```{r, eval = FALSE}
install.package("packageName")
remove.package("packageName")
```

## Different Ways to Time A Command

### Package `rbenchmark`

Function `benchmark` in this package compares the performance of functions (or 
assess performance of a single function, depending on the input) by replicating 
them multiple (100 by default) times and then reporting the average run time.

```{r}
library(rbenchmark)
f1 <- function(n) {
  s <- 0
  for (i in 1:n) {
    s <- s + i
  }
  s
}
f2 <- function(n) {
  sum(1:n)
}
f3 <- function(n) {
  n * (1 + n) / 2
}
benchmark(f1(1000), f2(1000), f3(1000), replications = 10000)
```

Package `microbenchmark` is able to assess run time of functions accurate to 
millisecond if necessary.

### Function `system.time()`

```{r}
count <- 0
system.time(for (i in 1:10000) count <- count + i ^ 2)
```

## File Manipulation

While function `system` enables programmers to use command line functionalities 
such as `cp` and `mv`, it no longer works in Windows environment because of the 
different syntax of system command followed. To avoid complication, it is 
better to use a group of R functions in `base` package to manipulate files and 
directories. The following example attempts to move a file `a.txt` in current 
directory into folder `../f2`, using function `file.rename`.

```{r eval = FALSE}
if (!file.exists("../f2")) {
  dir.create("../f2")
}
file.rename(from = "a.txt", to = "../f2/a.txt")
```

## Keyboard shortcuts

Action | macOS Shortcut | Windows Shortcut
:------------------------------------ | :------------ | :---------------
Run a single line or selection | `[Command] + [Return]` | `[Control] + [Return]` 
Source script | `[Command] + [Shift] + [S]` | `[Control] + [Shift] + [S]` 
Clear console | `[Control] + [L]` | `[Control] + [L]`

# General recommendation

## Avoid `for` loops

`for` loops are very slow in R. Replace them with functions from the `apply` 
family whenever possible, and use packages such as `plyr`, `dplyr` and 
`data.table` to manipulate data sets.

## Add Progress Bars

### For `for` loops

For programmes with long looping, always remember to output on screen which 
iteration the code is currently in, so as to access the performance of the code 
more conveniently.  

```{r, eval = FALSE}
total <- 20
# create progress bar
pb <- txtProgressBar(min = 0, max = total, style = 3)
for(i in 1:total){
  Sys.sleep(0.1)
  # update progress bar
  setTxtProgressBar(pb, i)
}
close(pb)
```

### For `apply` family

Use function `pb*pply` from package `pbapply` instead of `*pply`. 

## Avoid `factor` whenever possible

Operation done on `factor` may result output of other data types and hence 
cause error. Refer to the book [Advanced R](http://adv-r.had.co.nz/) for detail.

# Working with Different OS

macOS, Linux and Windows are the 3 major operating systems of current time. 
While macOS and Linux share many similarities, they are both quite different 
from Windows system in aspects such as file system and command line syntax.

## Choose the code to execute based on current OS environment

Function `Sys.info` returns system information including the name of OS in use.
Used together with control `switch`, this enables programmers to choose which 
OS-specific value to set or function to use. The example bellow shows how to 
set different working directories in different OS.

```{r eval = FALSE}
wd <- switch(Sys.info()[['sysname']],
             Windows = "/Users/mdcnyl/Google Drive/Active Projects/pig",
             Linux = "~/pig",
             Darwin = "~/Google Drive/Active Projects/pig") # macOS
setwd(wd)
```

# R Addins

## Addin Manager

```{r eval = FALSE}
devtools::install_github('rstudio/DT')
devtools::install_github("csgillespie/addinmanger")
```

When opened, a list of Addins available are shown, with Addins already installed 
highlighted in blue. To install new Addins, simply click on them to highlight 
them in blue as well. To remove existing Addins, click on them to remove the 
highlight. After clicking `Done`, current changes will be implemented and 
selected Addins will be installed or removed accordingly.

The list is not exhaustive, and the Addin cannot remove Addins that are not in 
the list.

## Colour Picker

```{r eval = FALSE}
install.packages("colourpicker")
```

This Addin allows users to choose colours in R by simple clicking. It allows 
users to select multiple colours at once.
