---
author: "Ning Yilin"
date: "Last update on `r Sys.Date()`"
title: "Mathematics and Basic Statistics"
---

```{r, include = FALSE}
# I personally prefer this style because it provides a floating table of 
# content:
'
output:
  html_document:
    theme: flatly
    highlight: zenburn
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
'
# But the knitted HTML file would be too large for Github to display. If you 
# are interested you can replace the current `output` chunk in the YAML code 
# above and try yourself.
```

<a name="top"></a>
<a style="display:scroll; position:fixed; bottom:5px; right:5px;" href="#top"><font size="-1">[top]</font></a>
<a style="display:scroll; position:fixed; bottom:5px; left:5px;" href="https://github.com/nyilin/My-R-help-sheets"><font size="-1">[Back to index]</font></a>

```{r, include=FALSE}
# global options
library(knitr)
opts_chunk$set(collapse = TRUE, echo = TRUE, eval = TRUE, warning = FALSE)
```

# Algebra

## Mathematical Operations

### Modulo

```{r}  
10 %% 3
10 %/% 3
```

# Linear Algebra

# Basic Statistics

## General Linear Model vs Generalised Linear Model

The following table is taken from [Wikipedia](https://en.wikipedia.org/wiki/Comparison_of_general_and_generalized_linear_models):

[]() | General linear model | Generalized linear model
:---------- | :----------------------------- | :--------------------------------
Typical estimation method	| Least squares, best linear unbiased prediction | Maximum likelihood or Bayesian
Special cases	| ANOVA, ANCOVA, MANOVA, MANCOVA, linear regression, mixed model | linear regression, logistic regression, Poisson regression, gamma regression
Function in R |	lm() | glm()

## Generating random variables from multivariate normal distribution

```{r, eval = FALSE}
# upper triangular matrix from cholescky decomposition
cholesckyMat <- chol(Sigma)
# n samples from standard normal distribution
xStd <- rnorm(n = n, mean = 0, sd = 1)
x <- muVec + cholesckyMat %*% xStd
```
