---
title: "Text"
author: "Ning Yilin"
date: "Last update on `r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    highlight: zenburn
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
---

```{r, include = FALSE}
# global options
library(knitr)
opts_chunk$set(collapse = TRUE, echo = TRUE, eval = TRUE)
```

# Cleaning and Manipulation

## Searching and Replacing

### Searching for Strings

### Replacing Strings

### Replacing Characters

## Combining

### Combining 

```{r}
x <- c("a b c", "d e", "f g h")
y <- c("1 2 3", "4 5", "6 7 8")
paste(x, sep = "_", collapse = "-")
paste(x, y, sep = "_", collapse = "-")
paste(x, y, sep = "_")
```

# Formating Output

## Capitalising Strings

The function below is taken from an answer on [this Stack Overflow page](http://stackoverflow.com/questions/24956546/capitalizing-letters-r-equivalent-of-excel-proper-function).

```{r}
x <- c('woRd Word', 'Word', 'word words')
gsub("(?<=\\b)([a-z])", "\\U\\1", tolower(x), perl = TRUE)
```
## `sprintf`  

Printing percentage:

```{r}
sprintf("%.2f%%", pi)
```

**Note**: to print percentage in `tex` file, use 
`paste0(round(pi * 100, 2), "\\%")` instead. This is implemented by function 
`printPercent` in package `myMISC`.
