---
title: "Data manipulation"
author: "Ning Yilin"
date: "Last update on `r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    highlight: zenburn
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE, eval = TRUE, echo = TRUE, warning = FALSE)
```

# Read Input from File

## Supported Formats

### `txt` and `csv`

### Excel Files

Function `read.xls` in package `gdata` can import `xls` and `xlsx` files into R 
through `perl`. An excel file will be converted to a `csv` file before 
importing, so the user must have the right to write in the system `tmp` folder. 
Apart from parameters in `read.xls` function, users could also use parameters 
from function `read.csv`.

In a Windows environment, `perl` is usually not pre-installed. In this case, 
user needs to manually download and install `perl`. But once `perl` is 
installed, the system will automatically update the directory to `perl.exe`, and
user can call `read.xls` data without having to specify the location of `perl`.

```{r, eval = FALSE}
library(gdata)
data <- read.xls(file = someInput, sheet = 1, header = TRUE, 
                 na.string = c("NA", "", " "), colClasses = "character")
```

Package `xlsx` also provides function to read in Excel files, but it has problem
with date and time variables.

### Stata Data Files

### SAS

### SPSS

## `read.table` vs `read.csv` vs `data.table::fread`

## Missingness

## Date and Time

# Output Data to File

## Writing table without column names

```{r}
#' Reading in file, preventing R from modifying its column names
#' @details When a column name in an input file contains space, R will join the 
#'   separate words with \code{.}. This function prevents R from doing this to 
#'   the column names.
#' @param file Path to the input file.
#' @param header Whether the input file has headers.
#' @param ... Other parameter that is accepted by built-in R function
#'   \code{read.csv}.
#' @return Returns a \code{data.frame} of the input data, with column names 
#'   exactly the same as in the original input file. 
readFileReserveHeader <- function(file, header = TRUE, ...) {
  if (header) {
    datain <- read.csv(file = file, header = FALSE, ...)
    cnames <- as.vector(sapply(datain[1, ], as.character))
    datain <- datain[-1, ]
    names(datain) <- cnames
    datain
  } else {
    read.csv(file = file, header = FALSE, ...)
  }
}
#' Writing data to a csv file, preventing R from modifying its column names
#' @details When a column name in a \code{data.frame} contains space, R will
#'   join the separate words with \code{.}. This function prevents R from doing
#'   this to the column names.
#' @param data The \code{data.frame} to write. Must have column names.
#' @param file Path to the input file.
writeFileReserveHeader <- function(data, file) {
  cnames <- names(data)
  data <- rbind(cnames, as.matrix(data))
  write.table(data, file = file, sep = ",", row.names = FALSE, 
              col.names = FALSE)
}
```

Users can specify strings to treat as `NA`. By default, texts will be read in 
as factors. To avoid this, use the `colClasses` parameter. This also avoids 
reading in date and time as strange numbers.

By default the input will be saved as a `data.frame`, even when there is only
one column in the data. 

```{r, eval = FALSE}
data <- read.csv(file = someInput, header = TRUE, na.string = c("NA", "", " "), 
                 colClasses = "character")
```

# Data Exploration

It is important to understand a new dataset before fitting any model to it or 
make any hypothesis based on it.

## Visualise

Use visualisation (link to the file).

## Missingness

When reading input from a `csv` file, empty columns and rows may also be 
included, which may cause unexpected errors. 

```{r, eval=FALSE}
# first of all, read in empty string as NA
table <- read.csv(file = file, na.string = c("NA", "", " "))

```

`complete.cases`
`anyNA`

Other codes used for missingess, e.g. `999`. Also depends on context. Read code 
book.

## Duplicates

```{r}
# find unique rows in a matrix
temp <- matrix(c(1, 1, 1, 2, 2, 2, 3, 3, 5), nrow = 3)
temp
unique(temp)
```

# Data Manipulation

## Subsetting

### Basic

- `which`
- `subset`

```{r}    
x <- c(1:20, 20:1)
y <- 1:40
tb <- cbind(x, y)
subset(x = tb, x > 11 & y < 14, select = x)
# package "dplyr" is more efficient for data subsetting and more manipulations.
```

#### Equality

```{r, eval=FALSE}
# do not use == to compare two double numbers
isTRUE(all.equal(x, y)) # not vectorised
```

#### Membership

```{r, eval=TRUE}
x <- c(1:20, 20:1)
y <- 1:40
tb <- cbind(x, y)
14 %in% x
14 %in% tb
match(x = 11, table = x)
match(x = 21, table = x)
match(x = 11, table = tb)
which(tb == 11)
```  

### `plyr`

### `ddply`

### `data.table`

## Sorting

### Single Vector

`order`
`sort`

### Multiple Vectors, Matrix, or `data.frame`

`order`
`sort.list`

```{r, eval=FALSE}
# sort data frame df first by column a, ascending, and then by column b, descending
df <- df[with(df, order(a, -b)), ]
```

## Merging

```{r}
A <- data.frame(x1 = c("a", "b", "c"), x2 = 1:3, x3 = rnorm(3))
A
B <- data.frame(y1 = c("a", "b"), y2 = 1:2, y3 = rnorm(2, 1, 1))
B
merge(A, B, by.x = c("x1", "x2"), by.y = c("y1", "y2"), all.x = TRUE)
```
