---
title: "Visualisation"
author: "Ning Yilin"
date: "Last update on `r Sys.Date()`"
---

```{r, include = FALSE}
# I personally prefer this style because it provides a floating table of 
# content:
'
output:
  html_document:
    theme: flatly
    highlight: zenburn
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
'
# But the knitted HTML file would be too large for Github to display. If you 
# are interested you can replace the current `output` chunk in the YAML code 
# above and try yourself.
```

<a name="top"></a>
<a style="display:scroll; position:fixed; bottom:5px; right:5px;" href="#top"><font size="-1">[top]</font></a>
<a style="display:scroll; position:fixed; bottom:5px; left:5px;" href="https://github.com/nyilin/My-R-help-sheets"><font size="-1">[Back to index]</font></a>

```{r, include = FALSE}
# global options
library(knitr)
opts_chunk$set(collapse = TRUE, echo = TRUE, eval = TRUE, cache = FALSE, 
               fig.width = 4, fig.height = 4)
```

# Choosing Colours for Visualisation

The choice of colour is important in visualisations. 
[ColorBrewer](http://colorbrewer2.org/) is a good website to choose colours for 
both qualitative and quantitative data.

Colours are specified using HEX values in R by default. There are also more than 
100 colours in R that have meaningful names. For example, colour `#FF8C00` is 
named `darkorange` in R. The most commonly used colours, such as black, red, 
green and blue, can also be referred to by numbers 1 to 4.

R also implements RGB colour specification with function `rgb`. Function 
`col2rgb` converts R colour to RBG values. For example, colour `darkorange` is 
`r paste(c("red", "green", "blue"), col2rgb("darkorange", alpha = FALSE), collapse = ", ", sep = " = ")` using RGB specification. These two functions are also 
members of the `grDevices` package.

With newer version of RStudio, we can also use Addin `Colour Picker` to choose 
colour conveniently. 

The ways of choosing colours mentioned above are convenient when only a few 
colours are needed. When colours are to be used to represent larger numbers of 
categories, or when we intend to use gradually changing colours to represent 
quantitative data, it is easier to use methods bellow.

There are several R functions that could generate colour for quantitative data.

### Rainbow

Function `rainbow` generates the seven preset colours in the rainbow. It is not 
very useful for quantitative data, as it only generates the 7 colours. If the 
length specified is more than 7, the 7 colours are repeated.

```{r}
len <- 100
plot(0, type = "n", xlim = c(0, len), ylim = c(0, len), xlab = "", ylab = "")
abline(h = 1:len, col = rainbow(len, alpha = 0.5))
```

## Gradually Changing Colour

Function `colorRampPalette` generates a vector of colours gradually changing 
from the first colour in the input to the second.

```{r}
plot(0, type = "n", xlim = c(0, len), ylim = c(0, len), xlab = "", ylab = "")
abline(h = 1:len, col = colorRampPalette(c("seashell", "steelblue"))(len))
```

## Heat colour

Function `heat.colors` generates a vector of heat colours, i.e. colour gradually 
changing from red to yellow.

```{r}
plot(0, type = "n", xlim = c(0, len), ylim = c(0, len), xlab = "", ylab = "")
abline(h = 1:len, col = heat.colors(len, alpha = 0.5))
```

## Other Pre-defined Colour Schemes

### `terrian`

```{r}
plot(0, type = "n", xlim = c(0, len), ylim = c(0, len), xlab = "", ylab = "")
abline(h = 1:len, col = terrain.colors(len, alpha = 0.5))
```

### `topo`

```{r}
plot(0, type = "n", xlim = c(0, len), ylim = c(0, len), xlab = "", ylab = "")
abline(h = 1:len, col = topo.colors(len, alpha = 0.5))
```

### `cm`

```{r}
plot(0, type = "n", xlim = c(0, len), ylim = c(0, len), xlab = "", ylab = "")
abline(h = 1:len, col = cm.colors(len, alpha = 0.5))
```

# Commonly Used Parameters

Parameters listed here are commonly used in plotting, and also applies to 
different plotting engines like `plot`, `grid` and `ggplot`.

## Point Characters

```{r, fig.height = 3, fig.width = 8}
plot(x = 1:25, y = numeric(25), pch = 1:25, main = "pch", xlab = "", ylab = "", 
     axes = FALSE, ylim = c(-0.5, 0.5))
text(x = 1:25, y = rep(-0.25, 25), labels = 1:25)
```

## Transparency

### Transparency for Contiguous Colours

Parameter `alpha` is available in all the contiguous colour functions, and 
controls the transparency of colours generated. The default value is 1 (opaque), 
and can be as low as 0 (invisible).

### Transparency for A Single Colour

Function `adjustcolor` in package `grDevices` (loaded by default) can be used 
to adjust the transparency of colour, ranging from 1 (opaque) to 0 (invisible).

```{r}
plot(0, type = "n", xlim = c(-1, 1), ylim = c(-1, 1), xlab = "", ylab = "")
polygon(c(-0.5, 0.5, 0.5, -0.5), c(-0.5, -0.5, 0.5, 0.5), col = "gray")
polygon(c(0, 1, 1, 0), c(0, 0, 1, 1), col = adjustcolor("blue", alpha.f = 0.5))
```

# Special Characters in Plots

R can plot special characters and expressions with the help of function 
`expression`.

```{r}
plot(0, type = "n", xlim = c(0, 1), ylim = c(0, 1), 
     xlab = expression(beta[1]), 
     ylab = expression(paste("Power for ", H[0], ": ", beta[1], " = 0")))
text(x = 0.1, y = 0.1, expression(theta[1] + theta[2]))
text(x = 0.2, y = 0.2, expression(32~degree~C))
text(x = 0.3, y = 0.3, labels = expression('' >= y))
# expression('' >= y >= z) does not work
```

# Using `plot`

## Axis and Labels  

Parameter `axes` can be used to remove axes and the box surrounding the plot. 
However, with the box removed, axes added to the plot look a bit wierd.

Function `axis` adds axes to plots. Note that parameter `las` controls the 
direction of axes labels, with `1` for horizontal and `3` for vertical. Other 
values are not implemented.

Function `mtext` can be used to add text right beside axes outside of the
plotting region. Parameter `las` works the same way in this function.

```{r}
plot(x = 1:5, y = 5:1, pch = 20, axes = FALSE, xlab = "x", ylab = "y")
axis(side = 1, at = c(1, 3, 5), las = 3)
axis(side = 2, at = c(2, 4), labels = paste0("y", c(2, 4)), las = 1)
mtext(text = "text by axis", side = 1, at = 4, las = 1)
```

To remove axes but keep the box and hence make the user-defined axes look 
better, use `xaxt` and `yaxt` instead. Note that `n` should be used instead of 
`FALSE`.

```{r}
plot(x = 1:5, y = 5:1, pch = 20, xaxt = "n", yaxt = "n", xlab = "x", ylab = "y")
axis(side = 1, at = c(1, 3, 5))
axis(side = 2, at = c(2, 4))
# text(x = xposition, y = yposition, text_to_add)
```	

## Text

Function `text` can be used to add text at any position in a plot.

```{r}
plot(x = 1:5, y = 5:1, pch = 20, xlab = "x", ylab = "y")
text(x = 3, y = 2, labels = "text here")
```

## Legend

```{r}
plot(x = 1:5, y = 5:1, pch = 20, col = 1:5, xlab = "x", ylab = "y")
legend("topright", legend = 1:5, col = 1:5, pch = rep(20, 5))
```

## Polygon 

```{r}
x <- 1:100
y <- rnorm(100)
plot(0, type = "n", xlim = c(1, 100), ylim = c(-4, 4), xlab = "x", ylab = "y")
polygon(x = c(40, 60, 60, 40), y = c(-2, -2, 2, 2), 
        col = "gray60", border = "gray60")
polygon(x = c(40, 60, 60, 40), y = c(-4, -4, -2, -2), 
        col = "gray50", border = "gray50")
lines(x = x, y = y)
```

## Two-Dimentional Kernel Density Plots

Contour plots are based on two-dimensional density functions, which can be 
calculated using function `kde2d` from package `MASS`.

Function `contour` plots contour plot for two-dementional kernel density.

```{r}
library(MASS)
attach(geyser)
den <- kde2d(geyser$waiting, geyser$duration, n = 50)
plot(x = geyser$waiting, y = geyser$duration, pch = 20, xlab = "waiting", 
     ylab = "duration")
contour(x = den, add = TRUE, col = "red")
```

## Scatter Plot with Contour Overlay

Function `pairs` plots pairwise scatter plots for all the variables, continuous 
or discrete, in a given matrix or data.frame. The functions for plots in the 
upper and lower pannels can be user-specified so that they display more 
information than mere scatter plots.

```{r}
pairs(geyser, 
      upper.panel = function(x, y) {
        points(x = x, y = y, pch = 20)
      }, 
      lower.panel = function(x, y) {
        r <- cor(x, y, method = "pearson")
        text(x = mean(range(x, na.rm = TRUE)), 
             y = mean(range(y, na.rm = TRUE)), 
             labels = sprintf("cor = %.3f", r))
      })
```
