---
title: "Time And Date"
author: "Ning Yilin"
date: "Last updated on `r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    highlight: zenburn
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
---

```{r, include = FALSE}
# global options
library(knitr)
opts_chunk$set(collapse = TRUE, echo = TRUE, eval = TRUE, cache = FALSE)
```

# Data Types

## `POSIXct`

- `typeof()`: `double`
- Number of seconds since `1970-01-01`

## `POSIXlt`

- `typeof()`: `list`
- `second`, `minute`, `hour`, `day of month`(1 - 31), `month of year`(0 - 11), 
`year`(since 1900), `day of week`(0 - 6), `day of year`(1 - 365), 
`daylight saving indices`

**When `print`ed, these two data types give the same result by default.**

Both data types can be compared using `>`, `<`, `==`, and are addable and 
substractable. But to calculate time difference between two time points, it is 
advisable to use function `difftime` instead:

```{r, echo = TRUE, eval = FALSE}
difftime(later, earlier, 
         units = c("auto", "secs", "mins", "hours", "days", "weeks"))
as.difftime(d, units)
```

# Type Conversion

## `strftime`

- Concerts time to string
- Must supply specific format to convert into

Units   | abbr | full | int
:------ | :--- | :--- | :-----------------------------------
weekday | `%a` | `%A` | `%u` (1 - 7) / `%w` (0 - 6)
month   | `%b` = `%h` | `%B` | `%m` (1 - 12)
year    | `%y` (00 - 99) | `%Y` | -
date    | - | `%d` / `%e` | `%j` (001 - 366)
hour    | - | - | `%H` (00 - 23) / `%I (0 - 12) + %p (AM, PM)`
minute  | - | - | `%M` (00 - 59)
second  | - | - | `%S` (00 - 59)

Shortcuts for certain formats:

Shortcut | Full format
:-------- | :----------
`%c`      | `%a %b %e %H:%M:%S %Y`
`%C`      | centry
`%D`      | `%m/%d/%y`
`%F`      | `%Y-%m-%d`
`%R`      | `%H:%M`
`%T`      | `%H:%M:%S`

## `strptime`

- Converts string to time
- Must supply specific format of input
- Output is of type `POSIXlt` by default
	  - `as.POSIXct(strptime(s, format))` converts the string `s` to `double`
	  - This is to make sure that the resulting time variables can be saved as 
	  elements in vectors and matrices
